\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-27 forwards-and-backwards.tex
alison comments

The purpose of the conditions that we have placed on the reproduction parameters is to 
ensure that the net per capita reproduction rate (before time scaling)
is order $1/\theta$. As remarked above, 
because of the non-local reproduction mechanism, it no longer suffices to assume that 
$r(x,\eta)\gamma(x,\eta)-\mu_\theta(x)$ is of order $1/\theta$.
Perhaps the simples example in which we can see this
is the case where $\gamma \equiv 1$ and $F \equiv 0$, so that $\mu_\theta = r$,
and $\eta = \delta_x$
(i.e., the population has all individuals at a single location),
so that $\smooth{r} \eta(y) = \rho_r(y)$.
In this case, the mean rate of change of the total population size
is $\int (r(y, \rho_r(y) ) - r(x, \rho_r(x))) q_\theta(x, dy)$;
the first condition of Lemma~\ref{lem:conditions on r} would ensure this is of order $1/\theta$.

@@

It is interesting to compare the stationary distribution we have obtained here
to the expression that we'd get by setting $\meanq = -\wavespeed$
and using Lemma~\ref{cor:stationary_dist},
i.e., by giving each offspring a mean displacement that offsets the motion of the wave.
In the Fisher-KPP and Allen-Cahn cases above, we get the same expressions,
but this is only the case because $r\equiv\gamma\equiv1$ in both.
For the PME with $\meanq = 1$ we have $\DG^* = \Delta + \grad$ 
and so the equation solved by the population density is
$$
    \partial_t \varphi = \Delta \varphi + 2 \varphi \grad \varphi + \varphi (1 - \varphi) ,
$$
which has a traveling wave solution of the same shape but moving at half the speed,
$\varphi(x,t) = w^{P}(x-t/2)$,
and a stationary distribution of the lineage relative to the wavefront of
$\pi(\xi) \propto e^{3\xi/2} (1 - e^{\xi/2})^3$.
% \comment{Details to back that last statement up, which we probably want to comment out?}
% (We can see that $w^P(x-t/2)$ is a traveling wave solution by using that
% $\partial_x w^P = -(1 - w^P)/2$;
% and then the generator of a linage in the moving frame is
% $\Lgen_{P'} = \varphi \partial_x^2 + (4 \partial_x \varphi + 1/2 + \varphi) \partial_x$,
% which has speed measure proportional to
% $\exp(\int^x 3(1/2 dx + d \log \varphi)) / \varphi$.)
% \begin{align*}
% m_{P'}(\xi)
%     &\propto \frac{1}{2(1-e^{\xi/2})} \exp\left(
%         \frac{3}{2} \int^\xi \left\{
%             1 - \frac{e^{x/2}}{1 - e^{x/2}}
%         \right\}dx
%     \right) .
% \end{align*}

@@

It is natural for applications to wonder about identifiability:
when can the observed quantities like population density
or certain summaries of lineage movement uniquely determine
the underlying demographic parameters?
Consider a deterministic,
continuous population generated by parameters $\gamma$, $r$, and $F$,
with $\meanq = 0$ and $\covq = I$.
Suppose it has a stationary profile $w(x)$, that must satisfy
$$
   r \Delta(\gamma w) + F w = 0 .
$$
It is easy to see that $w$ does not uniquely specify $\gamma$, $F$, and $r$:
let $\lambda(x)$ be a smooth, nonnegative function on $\IR^d$,
and let $r'(x, m) = \lambda(x) r(x, m)$ and $F'(x, m) = \lambda(x) F(x, m)$
(and, let $\gamma' = \gamma$).
Since $\mu = r \gamma - F/\theta$,
this corresponds to multiplying both establishment probabilities and death rates by $\lambda$.
Then the population with parameters $\gamma'$, $r'$, and $F'$
has the same stationary profile(s) as the original population.

@@

Now suppose instead that
Condition~\ref{control through gamma} of Lemma~\ref{lem:conditions on r} is satisfied.
First note that
    \begin{align}
	    \nonumber
    & |B_f^\theta| =
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x, \smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
 \label{eqn:q_bound_split}
	    &\qquad {} \leq
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y, \smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
\nonumber
	    &\qquad \qquad {}  +
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy) 
        \right| .
    \end{align}

@@

For the second claim, first define $\tau(t) = T(t)-t$.
Since if $\tau(t) \ge -(\epsilon^2 + t)/2$,
then $1/(\epsilon^2 + T(t)) \le 2/(\epsilon^2 + t)$,
while $\epsilon^2 + T(t) \ge \epsilon^2$ always,
partitioning over
$\{ \tau(t) \ge -(\epsilon^2 + t)/2 \}$ and its complement,
\begin{align}
   \| \psi^{\epsilon,x} \|_\infty
    &=
        \IE\left[ 
            \frac{1}{\big(2 \pi (\epsilon^2+T(t))\big)^{d/2}}
        \right] \nonumber
    \\ & \le
        \frac{C}{(\epsilon^2 + t)^{d/2}}
        + 
        \frac{C}{\epsilon^d}
        \IP\left\{
            \tau(t) < - (\epsilon^2 + t) / 2
        \right\}. \label{eqn:psi_infty_bound}
\end{align}
Now, observe that since $\IE[e^{-\Pi(\theta t)}] = \exp(-\theta t (1 - e^{-1}))$,
by Markov's inequality,
\begin{align}
    \IP\left\{
        \tau(t) < - \frac{\epsilon^2 + t}{2}
    \right\}
    &=
    \IP\left\{
        e^{-\Pi(\theta t)} > e^{-\theta (t - \epsilon^2) / 2}
    \right\}
\nonumber
\\
&\leq \frac{\IE[\exp\big(-\Pi(\theta t)\big)]}{\exp\big(-\theta(t-\epsilon^2)/2\big)}
\nonumber
    \\&=
    \frac{
        \exp(- \theta t(1 - e^{-1}))
    }{
        \exp(- \theta (t - \epsilon^2) / 2)
    }
\nonumber
    \\&=
    \exp\left\{ - \chi \theta t - \frac{\theta \epsilon^2}{2} \right\} ,
\label{bound for negative tau}
\end{align}
where $\chi = 1/2 - e^{-1} > 0$.
The second term in \eqref{eqn:psi_infty_bound} is therefore bounded by
\[
    C \left(1 + \frac{t}{\epsilon^2}\right)^{d/2} e^{-\chi \theta t - \epsilon^2 \theta / 2}
    \frac{1}{(\epsilon^2 + t)^{d/2}} .
\]
Now observe that the derivative (with respect to $t$) of 
$e^{-\chi \theta t - \epsilon^2 \theta / 2} (1 + t/\epsilon^2)^{d/2}$ 
is
\[
    \left( \frac{d}{2 \epsilon^2} - \left(1 + \frac{t}{\epsilon^2}\right) \chi \theta \right)
    \left( 1 + \frac{t}{\epsilon^2} \right)^{d/2 - 1} e^{-\chi \theta t - \epsilon^2 \theta / 2} ,
\]
which is negative if $\theta(\epsilon^2 + t) > d/2 \chi$.
At the maximum, $(\epsilon^2 + t) = d / (2\chi \theta)$,
so $(1 + t/\epsilon^2) = d /(2\chi \theta \epsilon^2)$,
and so this quantity is bounded uniformly over not only $t$ but also $\epsilon$
(since we've assumed that $\theta \epsilon^2$ is bounded below).
Therefore, the second term in \eqref{eqn:psi_infty_bound} 
is bounded by $C / (\epsilon^2 + t)^{d/2}$
(for a different constant $C$).
Substituting this into \eqref{eqn:psi_infty_bound} yields the result.
\end{proof}

@@

By the Intermediate Value Theorem, using $T_t$ to denote the heat semigroup,
there exists $s\in (0,1/\theta)$ such that
\begin{multline*}
\left|
\theta\int\big(\rho_\epsilon*f(y)-\rho_\epsilon*f(x)\big)
q_{\theta}(x,dy) \right|
=
\left|\theta\Big(T_{\epsilon^2+1/\theta}f(x)-T_{\epsilon^2}f(x)\Big)\right|
\\
=
\left|\partial_sT_{\epsilon^2+s}f(x)\right|=\left|T_{\epsilon^2+s}\Delta f(x)\right|
\leq\|\Delta f\|_\infty.
\end{multline*}
The first term in~\eqref{tightness estimate} is therefore bounded by
$$
    \| \Delta f \|_\infty |t-s| \sup_{0 \leq u \leq T } \langle 1, \eta_u^N \rangle .
$$
% We therefore have
% \begin{multline*}
%         \left|
%         \langle \rho_{\epsilon} * f, \eta_t^N \rangle
%         -
%         \langle \rho_{\epsilon} * f, \eta_s^N \rangle
%         \right|
%         \leq
%         \| \Delta f \|_\infty |t-s|
%         \sup_{0 \leq u \leq T } \langle 1, \eta_u^N \rangle
% \\
% +
% 	\int_s^t
% 	\big\langle \rho_\epsilon*|f|(x)\, \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
% \eta_u^N(dx)\big\rangle du +2\sup_{0\leq u\leq T}|\widehat{M}^N(f)_u|.
%     \end{multline*}
By the Burkholder-Davis-Gundy inequality,
$\IE[\sup_{0 \le u \le T} |\widehat{M}^N(f)_u|^2] \le C \IE[[\widehat{M}^N(f)]_u]$,
and so using
and Markov's inequality and
our moment bounds,
\begin{equation}
\label{martingale term to zero}
    \limsup_{N\to\infty}\IP\Big[
        2\sup_{0\leq u\leq T}|\widehat{M}^N(f)_u|
        >\frac{\nu}{3}
    \Big]
    \leq
    \limsup_{N\to\infty} \frac{36}{\nu^2} \IE\big[[\widehat{M}^N(f)]_T\big]
    \leq
    \limsup_{N\to\infty}\frac{36}{\nu^2}\frac{C\theta}{N\epsilon^d} 
    =0. 
\end{equation}
Now consider
\begin{multline}
	\label{square of F-integral}
\IE\Big[
	\Big(\int_s^t\big\langle\rho_\epsilon*|f|(x) \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du\Big)^2\Big]
\\
=
2\IE\Big[
	\int_s^t\big\langle\rho_\epsilon*|f|(x) \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle 
	\int_u^t\big\langle \rho_\epsilon*|f|(x)\big|F\big(\rho_\epsilon*\eta_r^N(x)\big)\big|,
\eta_r^N(dx)\big\rangle dr
du\Big].
\end{multline}
Since $F$ is polynomial, we use the approach
of Corollary~\ref{alternative form moment bounds}, the tower 
property, and 
Lemma~\ref{bounds on moments}, to bound this
in terms of sums of terms of the form
\[
\IE\Big[\int_s^t(t-u)\int\rho_\epsilon*|f|(x)\rho_\epsilon*\eta_u^N(x)^jdx
\int\rho_\epsilon*|f|(y)\rho_\epsilon*\eta_u^N(y)^kdydu\Big].
\]
Now observe that, again using Lemma~\ref{bounds on moments}, 
since $a^j b^k \le a^{j+k} b^{j+k}$,
\begin{multline*}
	\IE\left[	\int\int\rho_\epsilon*|f|(x)\rho_\epsilon*\eta_u^N(x)^j
	\rho_\epsilon*|f|(y)\rho_\epsilon*\eta_u^N(y)^k dx dy\right]
	\\
	\leq
	\IE\left[\int\int\|f\|_\infty\rho_\epsilon*\eta_u^N(x)^{j+k}
	\rho_\epsilon*|f|(y)
	dx dy
+
	\int\int
	\rho_\epsilon*|f|(x)
	\|f\|_\infty
	\rho_\epsilon*\eta_u^N(y)^{j+k}
	dx dy \right]
	\\ \leq C\int |f|(x)dx.
\end{multline*}
Thus the quantity~(\ref{square of F-integral}) 
is bounded by $C(t-s)^2$ for some constant $C$ which we can 
take to be independent of $s$, $t$ and $\epsilon$. Markov's inequality 
then gives
\[
\IP\Big[\|f\|_\infty\int_s^t
\big\langle \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du \geq \frac{\nu}{3}\Big]
\leq C\frac{(t-s)^2}{\nu^2}.
\]
A union bound gives that
\begin{equation}
\label{union bound}
\IP\Big[\max_i\|f\|_\infty\int_{t_{i-1}}^{t_i}
\big\langle \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du \geq \frac{\nu}{3}\Big]
\leq C\frac{T\delta}{\nu^2}.
\end{equation}
Now using Markov's inequality, we can choose $K$ so that
\[
\IP\Big[\|\Delta f\|_\infty\, \sup_{0\leq t\leq T} \langle 1,\eta_t^N\rangle 
>K\Big]<\frac{\nu}{3}, 
\]
and so choosing $\delta$ so that $K\delta<\nu/3$ in this 
expression and $C\delta<\nu^3/3T$ in~(\ref{union bound}),
combining with~(\ref{martingale term to zero}), the result follows.
\end{proof}

@@

\begin{proof}
    Still using the notation of Lemma~\ref{PsiBoundHS},
%the three terms on the right hand side of 
%equation~(\ref{heat kernel estimate})
%    correspond to bounds obtained by partitioning
 we partition into three events according to the value of $\tau(t)$.
    Let $A_1 = \{ \tau(t) < - (\epsilon^2 + t)/2 \}$,
    $A_2 = \{ \tau(t) > 2(\epsilon^2 + t) \}$,
    and $A_3$ the remaining event, $\{ - (\epsilon^2 + t)/2 \le \tau(t) \le 2(\epsilon^2 + t) \}$.
    Then,
    \begin{align*}
        \left|
            \psi_t^{\epsilon, x}(y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right|
        &=
        \left|
        \IE\left[
            p_{\epsilon^2 + t + \tau(t)}(x, y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right]
        \right|
        \\&\le
        \IE\left[
            (1_{A_1} + 1_{A_2} + 1_{A_3})
            \left|
            p_{\epsilon^2 + t + \tau(t)}(x, y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right|
        \right] .
    \end{align*}
    For the first term, note that if $a < b$ then
    \begin{align*}
        |p_a(x, y) - p_b(x, y)| 
        &=
        \frac{1}{(2\pi)^{d/2}}
        \left|
            \frac{1}{a^{d/2}}
            e^{-\|x - y\|^2 / 2a}
            -
            \frac{1}{b^{d/2}}
            e^{-\|x - y\|^2 / 2b}
        \right|
        \\ &=
        \frac{1}{(2\pi a^2)^{d/2}}
            e^{-\|x - y\|^2 / 2b}
        \left|
            e^{-\|x - y\|^2 \left(\frac{1}{2a} - \frac{1}{2b}\right)}
            -
            \left(\frac{a}{b}\right)^{d/2}
        \right|
        \\ &\le
        C \left(\frac{b}{a}\right)^{d/2}
        p_b(x, y) ,
    \end{align*}
    where the inequality follows because both terms under the absolute value 
are less than 1.
Since, on the event $A_1$, $\tau(t)<0$, we can  
apply this with $a = \epsilon^2 + t + \tau(t)$ and $b = \epsilon^2 + t$,
and, using the bound~(\ref{bound for negative tau}),
    \begin{align*}
        \IE\left[
            1_{A_1} |p_a(x, y) - p_b(x, y)| 
        \right]
        & \le
            C \left(\frac{\epsilon^2 + t}{\epsilon^2}\right)^{d/2}
            p_{\epsilon^2 + t}(x, y) 
            \IP\left\{ \tau(t) < - \frac{\epsilon^2 + t}{2} \right\}
        \\ & \le
            C \frac{1}{\epsilon^d}
            \IP\left\{ \tau(t) < - \frac{\epsilon^2 + t}{2} \right\}
        \\ & \le
        C \exp\left(-\frac{\theta \epsilon^2}{2} \right) .
    \end{align*}

>>>>

The purpose of the conditions that we have placed on the reproduction parameters is to 
ensure that the net per capita reproduction rate (before time scaling)
is order $1/\theta$. As remarked above, 
because of the non-local reproduction mechanism, it no longer suffices to assume that 
$r(x,\eta)\gamma(x,\eta)-\mu_\theta(x)$ is of order $1/\theta$.
Perhaps the simplest example in which we can see this
is the case where $\gamma \equiv 1$ and $F \equiv 0$, so that $\mu_\theta = r$,
and $\eta = \delta_x$
(i.e., the population has all individuals at a single location),
so that $\smooth{r} \eta(y) = \rho_r(y)$.
In this case, the mean rate of change of the total population size
is $\int (r(y, \rho_r(y) ) - r(x, \rho_r(x))) q_\theta(x, dy)$;
the first condition of Lemma~\ref{lem:conditions on r} would ensure this is of order $1/\theta$.

@@

It is interesting to compare the stationary distribution we have obtained here
to the expression that we'd get by setting $\meanq = -\wavespeed$
and using Corollary~\ref{cor:stationary_dist},
i.e., by giving each offspring a mean displacement that offsets the motion of the wave.
In the Fisher-KPP and Allen-Cahn cases above, we get the same expressions,
but this is only the case because $r\equiv\gamma\equiv1$ in both.
For the PME with $\meanq = 1$ we have $\DG^* = \Delta + \grad$ 
and so the equation solved by the population density is
$$
    \partial_t \varphi = \Delta \varphi + 2 \varphi \grad \varphi + \varphi (1 - \varphi) ,
$$
which has a traveling wave solution of the same shape but moving at half the speed,
$\varphi(x,t) = w^{P}(x-t/2)$,
and a stationary distribution of the lineage relative to the wavefront of
$\pi(\xi) \propto e^{3\xi/2} (1 - e^{\xi/2})^3$.
% \comment{Details to back that last statement up, which we probably want to comment out?}
% (We can see that $w^P(x-t/2)$ is a traveling wave solution by using that
% $\partial_x w^P = -(1 - w^P)/2$;
% and then the generator of a linage in the moving frame is
% $\Lgen_{P'} = \varphi \partial_x^2 + (4 \partial_x \varphi + 1/2 + \varphi) \partial_x$,
% which has speed measure proportional to
% $\exp(\int^x 3(1/2 dx + d \log \varphi)) / \varphi$.)
% \begin{align*}
% m_{P'}(\xi)
%     &\propto \frac{1}{2(1-e^{\xi/2})} \exp\left(
%         \frac{3}{2} \int^\xi \left\{
%             1 - \frac{e^{x/2}}{1 - e^{x/2}}
%         \right\}dx
%     \right) .
% \end{align*}

@@

It is natural for applications to wonder about identifiability:
when can the observed quantities like population density
or certain summaries of lineage movement uniquely determine
the underlying demographic parameters?
Consider a deterministic,
continuous population generated by parameters $\gamma$, $r$, and $F$,
with $\meanq = 0$ and $\covq = I$.
Suppose it has a stationary profile $w(x)$, that must satisfy
$$
   r \Delta(\gamma w) + F w = 0 .
$$
It is easy to see that $w$ does not uniquely specify $\gamma$, $F$, and $r$:
let $\lambda(x)$ be a smooth, nonnegative function on $\IR^d$,
\comment{TODO Should we worry that we used prime for differentiation before?}
and let $r'(x, m) = \lambda(x) r(x, m)$ and $F'(x, m) = \lambda(x) F(x, m)$
(and, let $\gamma' = \gamma$).
Since $\mu = r \gamma - F/\theta$,
this corresponds to multiplying both establishment probabilities and death rates by $\lambda$.
Then the population with parameters $\gamma'$, $r'$, and $F'$
has the same stationary profile(s) as the original population.

@@

\comment{TODO I think we need to make the notation for $r(y, \rho_r*\eta(y))$ 
consistent with what we have (now) used elsewhere, namely $r(y,\eta)$?}
Now suppose instead that
Condition~\ref{control through gamma} of Lemma~\ref{lem:conditions on r} is satisfied.
First note that
    \begin{align}
	    \nonumber
    & |B_f^\theta| =
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x, \smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
 \label{eqn:q_bound_split}
	    &\qquad {} \leq
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(y) r\big(y, \smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                \big\}
            q_{\theta}(x,dy)
        \right|
    \\ 
\nonumber
	    &\qquad \qquad {}  +
        \left| \theta 
            \int_{\IR^n} 
                \big\{
                    f(x) r\big(x,\smooth{r}\eta(y)\big)
                    -
                    f(x) r\big(x,\smooth{r}\eta(x)\big)
                \big\}
            q_{\theta}(x,dy) 
        \right| .
    \end{align}

@@

For the second claim, first define $\tau(t) = T(t)-t$.
Since if $\tau(t) \ge -(\epsilon^2 + t)/2$,
then $1/(\epsilon^2 + T(t)) \le 2/(\epsilon^2 + t)$,
while $\epsilon^2 + T(t) \ge \epsilon^2$ always,
partitioning over
$\{ \tau(t) \ge -(\epsilon^2 + t)/2 \}$ and its complement,
\begin{align}
   \| \psi^{\epsilon,x} \|_\infty
    &=
        \IE\left[ 
            \frac{1}{\big(2 \pi (\epsilon^2+T(t))\big)^{d/2}}
        \right] \nonumber
    \\ & \le
        \frac{C}{(\epsilon^2 + t)^{d/2}}
        + 
        \frac{C}{\epsilon^d}
        \IP\left\{
            \tau(t) < - (\epsilon^2 + t) / 2
        \right\}. \label{eqn:psi_infty_bound}
\end{align}
Now, observe that since $\IE[e^{-\Pi(\theta t)}] = \exp(-\theta t (1 - e^{-1}))$,
by Markov's inequality,
\begin{align}
    \IP\left\{
        \tau(t) < - \frac{\epsilon^2 + t}{2}
    \right\}
    &=
    \IP\left\{
        e^{-\Pi(\theta t)} > e^{-\theta (t - \epsilon^2) / 2}
    \right\}
\nonumber
\\
&\leq \frac{\IE[\exp\big(-\Pi(\theta t)\big)]}{\exp\big(-\theta(t-\epsilon^2)/2\big)}
\nonumber
    \\&=
    \frac{
        \exp(- \theta t(1 - e^{-1}))
    }{
        \exp(- \theta (t - \epsilon^2) / 2)
    }
\nonumber
    \\&=
    \exp\left\{ - \chi \theta t - \frac{\theta \epsilon^2}{2} \right\} ,
\label{bound for negative tau}
\end{align}
where $\chi = 1/2 - e^{-1} > 0$.
The second term in \eqref{eqn:psi_infty_bound} is therefore bounded by
\[
    C \left(1 + \frac{t}{\epsilon^2}\right)^{d/2} e^{-\chi \theta t - \epsilon^2 \theta / 2}
    \frac{1}{(\epsilon^2 + t)^{d/2}} .
\]
Now observe that the derivative (with respect to $t$) of 
$e^{-\chi \theta t - \epsilon^2 \theta / 2} (1 + t/\epsilon^2)^{d/2}$ 
is
\[
    \left( \frac{d}{2 \epsilon^2} - \left(1 + \frac{t}{\epsilon^2}\right) \chi \theta \right)
    \left( 1 + \frac{t}{\epsilon^2} \right)^{d/2 - 1} e^{-\chi \theta t - \epsilon^2 \theta / 2} ,
\]
which is negative if $\theta(\epsilon^2 + t) > d/2 \chi$.
At the maximum, $(\epsilon^2 + t) = d / (2\chi \theta)$,
so $(1 + t/\epsilon^2) = d /(2\chi \theta \epsilon^2)$,
and so this quantity is bounded uniformly over not only $t$ but also $\epsilon$
(since we've assumed that $\theta \epsilon^2$ is bounded below).
Therefore, the second term in \eqref{eqn:psi_infty_bound} 
is bounded by $C / (\epsilon^2 + t)^{d/2}$
(for a different constant $C$).
Substituting this into \eqref{eqn:psi_infty_bound} yields the result.
\end{proof}
\comment{ALISON: Actually the last calculation gives a much better bound which we will
need later. I'll sort this out tomorrow.}

@@

By the Intermediate Value Theorem, using $T_t$ to denote the heat semigroup,
there exists $s\in (0,1/\theta)$ such that
\begin{multline*}
\left|
\theta\int\big(\rho_\epsilon*f(y)-\rho_\epsilon*f(x)\big)
q_{\theta}(x,dy) \right|
=
\left|\theta\Big(T_{\epsilon^2+1/\theta}f(x)-T_{\epsilon^2}f(x)\Big)\right|
\\
=
\left|\partial_sT_{\epsilon^2+s}f(x)\right|=\left|T_{\epsilon^2+s}\Delta f(x)\right|
\leq\|\Delta f\|_\infty.
\end{multline*}
The first term in~\eqref{tightness estimate} is therefore bounded by
$$
    \| \Delta f \|_\infty |t-s| \sup_{0 \leq u \leq T } \langle 1, \eta_u^N \rangle .
$$

By the Burkholder-Davis-Gundy inequality,
$\IE[\sup_{0 \le u \le T} |\widehat{M}^N(f)_u|^2] \le C \IE[[\widehat{M}^N(f)]_u]$,
and so using
and Markov's inequality and
our moment bounds,
\begin{equation}
\label{martingale term to zero}
    \limsup_{N\to\infty}\IP\Big[
        2\sup_{0\leq u\leq T}|\widehat{M}^N(f)_u|
        >\frac{\nu}{3}
    \Big]
    \leq
    \limsup_{N\to\infty} \frac{36}{\nu^2} \IE\big[[\widehat{M}^N(f)]_T\big]
    \leq
    \limsup_{N\to\infty}\frac{36}{\nu^2}\frac{C\theta}{N\epsilon^d} 
    =0. 
\end{equation}
Now consider
\begin{multline}
	\label{square of F-integral}
\IE\Big[
	\Big(\int_s^t\big\langle\rho_\epsilon*|f|(x) \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du\Big)^2\Big]
\\
=
2\IE\Big[
	\int_s^t\big\langle\rho_\epsilon*|f|(x) \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle 
	\int_u^t\big\langle \rho_\epsilon*|f|(x)\big|F\big(\rho_\epsilon*\eta_r^N(x)\big)\big|,
\eta_r^N(dx)\big\rangle dr
du\Big].
\end{multline}
Since $F$ is polynomial, we use the approach
of Corollary~\ref{alternative form moment bounds}, the tower 
property, and 
Lemma~\ref{bounds on moments}, to bound this
in terms of sums of terms of the form
\[
\IE\Big[\int_s^t(t-u)\int\rho_\epsilon*|f|(x)\rho_\epsilon*\eta_u^N(x)^jdx
\int\rho_\epsilon*|f|(y)\rho_\epsilon*\eta_u^N(y)^kdydu\Big].
\]
Now observe that, again using Lemma~\ref{bounds on moments}, 
since $a^j b^k \le a^{j+k} b^{j+k}$,
\begin{multline*}
	\IE\left[	\int\int\rho_\epsilon*|f|(x)\rho_\epsilon*\eta_u^N(x)^j
	\rho_\epsilon*|f|(y)\rho_\epsilon*\eta_u^N(y)^k dx dy\right]
	\\
	\leq
	\IE\left[\int\int\|f\|_\infty\rho_\epsilon*\eta_u^N(x)^{j+k}
	\rho_\epsilon*|f|(y)
	dx dy
+
	\int\int
	\rho_\epsilon*|f|(x)
	\|f\|_\infty
	\rho_\epsilon*\eta_u^N(y)^{j+k}
	dx dy \right]
	\\ \leq C\int |f|(x)dx.
\end{multline*}
Thus the quantity~(\ref{square of F-integral}) 
is bounded by $C(t-s)^2$ for some constant $C$ which we can 
take to be independent of $s$, $t$ and $\epsilon$. Markov's inequality 
then gives
\[
\IP\Big[\|f\|_\infty\int_s^t
\big\langle \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du \geq \frac{\nu}{3}\Big]
\leq C\frac{(t-s)^2}{\nu^2}.
\]
A union bound gives that
\begin{equation}
\label{union bound}
\IP\Big[\max_i\|f\|_\infty\int_{t_{i-1}}^{t_i}
\big\langle \big|F\big(\rho_\epsilon*\eta_u^N(x)\big)\big|,
\eta_u^N(dx)\big\rangle du \geq \frac{\nu}{3}\Big]
\leq C\frac{T\delta}{\nu^2}.
\end{equation}
Now using Markov's inequality, we can choose $K$ so that
\[
\IP\Big[\|\Delta f\|_\infty\, \sup_{0\leq t\leq T} \langle 1,\eta_t^N\rangle 
>K\Big]<\frac{\nu}{3}, 
\]
and so choosing $\delta$ so that $K\delta<\nu/3$ in this 
expression and $C\delta<\nu^3/3T$ in~(\ref{union bound}),
combining with~(\ref{martingale term to zero}), the result follows.
\end{proof}

@@

\begin{proof}
    Still using the notation of Lemma~\ref{PsiBoundHS},
%the three terms on the right hand side of 
%equation~(\ref{heat kernel estimate})
%    correspond to bounds obtained by partitioning
 we partition into three events according to the value of $\tau(t)$.
    Let $A_1 = \{ \tau(t) < - (\epsilon^2 + t)/2 \}$,
    $A_2 = \{ \tau(t) > 2(\epsilon^2 + t) \}$,
    and $A_3$ the remaining event, $\{ - (\epsilon^2 + t)/2 \le \tau(t) \le 2(\epsilon^2 + t) \}$.
    Then,
    \begin{align*}
        \left|
            \psi_t^{\epsilon, x}(y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right|
        &=
        \left|
        \IE\left[
            p_{\epsilon^2 + t + \tau(t)}(x, y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right]
        \right|
        \\&\le
        \IE\left[
            (1_{A_1} + 1_{A_2} + 1_{A_3})
            \left|
            p_{\epsilon^2 + t + \tau(t)}(x, y)
            -
            p_{\epsilon^2 + t}(x, y)
        \right|
        \right] .
    \end{align*}
    For the first term, note that if $a < b$ then
    \begin{align*}
        |p_a(x, y) - p_b(x, y)| 
        &=
        \frac{1}{(2\pi)^{d/2}}
        \left|
            \frac{1}{a^{d/2}}
            e^{-\|x - y\|^2 / 2a}
            -
            \frac{1}{b^{d/2}}
            e^{-\|x - y\|^2 / 2b}
        \right|
        \\ &=
        \frac{1}{(2\pi a^2)^{d/2}}
            e^{-\|x - y\|^2 / 2b}
        \left|
            e^{-\|x - y\|^2 \left(\frac{1}{2a} - \frac{1}{2b}\right)}
            -
            \left(\frac{a}{b}\right)^{d/2}
        \right|
        \\ &\le
        C \left(\frac{b}{a}\right)^{d/2}
        p_b(x, y) ,
    \end{align*}
    where the inequality follows because both terms under the absolute value 
are less than 1.
Since, on the event $A_1$, $\tau(t)<0$, we can  
apply this with $a = \epsilon^2 + t + \tau(t)$ and $b = \epsilon^2 + t$,
and, using the bound~(\ref{bound for negative tau}),
\comment{ALISON: It's actually using the bound derived just below (7.4) which is 
much stronger than stated there}
    \begin{align*}
        \IE\left[
            1_{A_1} |p_a(x, y) - p_b(x, y)| 
        \right]
        & \le
            C \left(\frac{\epsilon^2 + t}{\epsilon^2}\right)^{d/2}
            p_{\epsilon^2 + t}(x, y) 
            \IP\left\{ \tau(t) < - \frac{\epsilon^2 + t}{2} \right\}
        \\ & \le
            C \frac{1}{\epsilon^d}
            \IP\left\{ \tau(t) < - \frac{\epsilon^2 + t}{2} \right\}
        \\ & \le
        C \exp\left(-\frac{\theta \epsilon^2}{2} \right) .
    \end{align*}
